---
title: "StormDataAnalysis"
author: "Sebastian Cruz"
date: "April 22, 2015"
output: html_document
---

##Summary
This is my summary text.

##Data Processing
*Load Required Libraries*
```{r, echo=TRUE, message=FALSE}
library(ggplot2)
library(car) #recode() 

```

*Download and Import*
```{r, echo=TRUE, message=FALSE}
# Set Working directory
# Edit as necessary
setwd("~/GitHub/RepData_PeerAssessment2")

# Import Data
data.raw <- read.csv(bzfile("repdata-data-StormData.csv.bz2")) #Can read .bz2 files directly with bzfile()
```

*Standardize Values and Units*
Property Damage (PROPDMG) and Crop Damage (CROPDMG) are the estimated financial damage caused.  The value magnitudes are determined by the characters in the fields PROPDMGEXP and CROPDMGEXP according to the [data documentation] (https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) provided, see Section 2.7 Damage.
Value	|	Mapping
-----	-	-------
-	|	0
?	|	0
+	|	0
B	|	Billions
h	|	Hundreds
H	|	Hundreds
K	|	Thousands
M	|	Millions
m	|	Millions
0-9 	|	Exponent
```{r, echo=TRUE, message=FALSE}


```

