---
title: "StormDataAnalysis"
author: "Sebastian Cruz"
date: "April 22, 2015"
output: pdf_document
---

###Synopsis
For the United States between 1950 and 2011 and using data from the [NOAA Storm Database](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) I isolated the weather events responsible for the most Deaths and Injuries and also the events responsible for the greatest economic harm.

In terms of Death and Injuries, Tornados were far and away the biggest cause of totalled death and injuries. They were responsible for 96,979 deaths and injuries or 62% of the total.

In terms of Economic Damage, Flooding is the top weather event.  Flooding was responsible for over $150B in damages or 31% of the total for the dataset.

Despite the dangers of Tornados and Floods pose to residence of the Unitied States, there is not a single law making these events illegal. Based on this analysis I would strongly encourage the legislature to outlaw Torandos and Floods to save lives and economic resources.

###Data Processing
*Load Required Libraries*
```{r, echo=TRUE, message=FALSE}
library(ggplot2)
library(car) #recode()
library(dplyr)
```

*Import Data*
```{r, echo=TRUE, message=FALSE, cache=TRUE}
# Set Working directory
# Edit as necessary
setwd("~/GitHub/RepData_PeerAssessment2")

# Import Data
data.raw = read.csv(bzfile("repdata-data-StormData.csv.bz2")) #Can read .bz2 files directly with bzfile()
```

*Isolate Required Columns*
```{r, echo=TRUE, message=FALSE, cache=TRUE}
# Isolate Required Columns
data.analysis = data.raw[ ,c('EVTYPE', 'FATALITIES', 'INJURIES', 'PROPDMG', 'CROPDMG', 'PROPDMGEXP', 'CROPDMGEXP')]
head(data.analysis)
```

*Standardize Values and Magnitude*
Property Damage (PROPDMG) and Crop Damage (CROPDMG) are the estimated financial damage caused.  The value magnitudes are determined by the characters in the fields PROPDMGEXP and CROPDMGEXP according to the [data documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) provided, see Section 2.7 Damage. I am assuming a mapping value of 1 for the non-alphanumeric values and assuming that numerical values are the number of significant digits used in scientific notation.

Value | Mapping
----- | -------
'' | 1
- | 1
? | 1
+ | 1
B | Billions
h | Hundreds
H | Hundreds
K | Thousands
M | Millions
m | Millions
0-9 | Significant Digits

```{r, echo=TRUE, message=FALSE, cache=TRUE}
#Recode PROPDMGEXP values
data.analysis$PROPDMGEXP = as.numeric(
  recode(data.analysis$PROPDMGEXP, "
    c('-','?','+','') = 1;
    c('h','H') = 1e+02;
    c('k','K') = 1e+03;
    c('m','M') = 1e+06;
    c('b','B') = 1e+09;
    '0' = 1; 
    '1' = 10;
    '2' = 1e+02; 
    '3' = 1e+03; 
    '4' = 1e+04; 
    '5' = 1e+05;
    '6' = 1e+06; 
    '7' = 1e+07; 
    '8' = 1e+08;
    ", as.factor.result = FALSE
  )
)

#Update PROPDMG values with new mumeric PROPDMGEXP values
data.analysis$PROPDMG = data.analysis$PROPDMG * data.analysis$PROPDMGEXP

#Recode CROPDMGEXP values
data.analysis$CROPDMGEXP = as.numeric(
  recode(data.analysis$CROPDMGEXP, "
    c('-','?','+','') = 1;
    c('h','H') = 1e+02;
    c('k','K') = 1e+03;
    c('m','M') = 1e+06;
    c('b','B') = 1e+09;
    '0' = 1; 
    '1' = 10;
    '2' = 1e+02; 
    '3' = 1e+03; 
    '4' = 1e+04; 
    '5' = 1e+05;
    '6' = 1e+06; 
    '7' = 1e+07; 
    '8' = 1e+08;
    ", as.factor.result = FALSE
  )
)

#Update CROPDMG values with new mumeric CROPDMGEXP values
data.analysis$CROPDMG = data.analysis$CROPDMG * data.analysis$CROPDMGEXP
head(data.analysis)
```

*Add Totals for Damage and Health*
```{r, echo=TRUE, message=FALSE}
#Add PROPDMG + CROPDMG as TOTALDMG
data.analysis$TOTALDMG = data.analysis$PROPDMG + data.analysis$CROPDMG

#Add FATALITIES + INJURIES as TOTALDEATHINJ
data.analysis$TOTALDEATHINJ = data.analysis$FATALITIES + data.analysis$INJURIES
head(data.analysis)
```

*Summarize Health and Damage*
The summerized datasets will be used in the analysis below
```{r, echo=TRUE, message=FALSE}
#Create .health dataset, Remove EXP columns for clean summerization
data.analysis.health = data.analysis[ ,c('EVTYPE','FATALITIES','INJURIES','TOTALDEATHINJ')]

#Create .damage dataset, Remove EXP columns for clean summerization
data.analysis.damage = data.analysis[ ,c('EVTYPE','PROPDMG','CROPDMG','TOTALDMG')]

#Summerize Health using dplyr
data.analysis.health.sum = 
  data.analysis.health %>%
  group_by(EVTYPE) %>%
  summarise_each(funs(sum)) %>%
  arrange(desc(TOTALDEATHINJ))

head(data.analysis.health.sum)

#Summerize Damage using dplyr
data.analysis.damage.sum = 
  data.analysis.damage %>%
  group_by(EVTYPE) %>%
  summarise_each(funs(sum)) %>%
  arrange(desc(TOTALDMG))

head(data.analysis.damage.sum)
```


###Results 
####1. Across the United States, which types of events (as indicated in the `EVTYPE` variable) are most harmful with respect to population health?
To answer this I utilized the `data.analysis.health.sum` summary table that sums the deaths and injuries by `EVTYPE` ordered descending. Additionally I isolated the top 10 causes of Total Death & Injuries and calculated the proportion they make as a percentage of the whole.  

```{r, echo=TRUE, message=FALSE}
#Top 10 Events
data.analysis.health.sum[1:10, ]

#Calc Specific Sums and %
AllUp.DeathInj = sum(data.analysis.health.sum$TOTALDEATHINJ)
Top10.DeathInj = sum(data.analysis.health.sum[1:10,4])
PrctTop10.DeathInj = Top10.DeathInj/AllUp.DeathInj
TopEVTYPE.DeathInj = sum(data.analysis.health.sum[1,4])
PrctTopEVTYPE.DeathInj = TopEVTYPE.DeathInj/AllUp.DeathInj
```

Tornados cause the vast majority of death and injuries in the United States between 1950 and 2011. They were responsible for `r TopEVTYPE.DeathInj` or `r round(PrctTopEVTYPE.DeathInj * 100, digits=1)`% of all Deaths and Injuries.

```{r, message=FALSE}
ggplot(data.analysis.health.sum[1:10, ]) +
  geom_bar(aes(reorder(EVTYPE, TOTALDEATHINJ, sum), y=TOTALDEATHINJ), width=0.5, stat="identity") +
  coord_flip() +
  ggtitle("Top 10 Weather Events by Total Deaths & Injuries") +
  ylab("Total Deaths & Injuries") +
  xlab("Weather Events")
```

####2. Across the United States, which types of events have the greatest economic consequences?
To answer this I utilized the `data.analysis.damage.sum` summary table that sums the total damage (`PROPDMG` + `CROPDMG` as `TOTALDMG`) by `EVTYPE` ordered descending.  Additionally I isolated the top 10 causes of Total Damage and calculated the proportion they make as a percentage of the whole.

```{r, message=FALSE}
#Top 10 Events
data.analysis.damage.sum[1:10, ]

#Calc Specific Sums and %
AllUp.Damage = sum(data.analysis.damage.sum$TOTALDMG)
Top10.Damage = sum(data.analysis.damage.sum[1:10,4])
PrctTop10.Damage = Top10.Damage/AllUp.Damage
TopEVTYPE.Damage = sum(data.analysis.damage.sum[1,4])
PrctTopEVTYPE.Damage = TopEVTYPE.Damage/AllUp.Damage
```

Flooding is responsible for the vast majority of Total Damage in the United States between 1950 and 2011.  Floods were responsible for $`r round(TopEVTYPE.Damage/1e+09, digits=1)`B in Total Damages or `r round(PrctTopEVTYPE.Damage * 100, digits=1)`% of all Damage Types.

```{r, message=FALSE}
ggplot(data.analysis.damage.sum[1:10, ]) +
  geom_bar(aes(reorder(EVTYPE, TOTALDMG, sum), y=TOTALDMG/1e+09), width=0.5, stat="identity") +
  coord_flip() +
  ggtitle("Top 10 Weather Events by Total Damage") +
  ylab("Total Damage in $ Billions") +
  xlab("Weather Events")
```


